---
title: "Analyze XP_028889033 family evolution and adhesin properties"
author: "Bin He"
date: "7/16/2020"
output: 
  html_notebook:
    toc: yes
    toc_float: yes
    toc_depth: 5
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r load_libraries, echo = FALSE}
if (!requireNamespace("BiocManager", quietly = TRUE))
    install.packages("BiocManager")
if (!requireNamespace("treeio", quietly = TRUE))
    BiocManager::install("treeio")
if (!requireNamespace("ggtree", quietly = TRUE))
    BiocManager::install("ggtree")
if (!requireNamespace("reticulate", quietly = TRUE))
    install.packages("reticulate")
library(tidyverse)
library(reticulate)
use_python("")
```

## Goal

Analyze the gene tree and the adhesin properties of the XP_028889033 homologs (putative adhesin in _C. auris_)

## Approach
1. Load gene tree
1. Parse various computational prediction results and assemble a master table with one row for each homolog
1. Plot the data

## Build datasets
First get the basic information about the 110 sequences in this study. I decide to write a simple Python script to extract such info.
```{bash}
# edit the PYTHON path below to match your local system
~/sw/miniconda3/bin/python extract_seq_info.py
```

### Sequence info
Load in the sequence information.
```{r load_seq_info}
seqInfo <- read_tsv("XP_028889033_homologs.tsv", comment = "#", col_types = "ccci") %>% 
  mutate(species = factor(species, levels = c("Cduobushaemulonis","Cpseudohaemulonis","Chaemuloni","Cauris","Clusitaniae","Dhansenii","Cparapsilosis","Lelongisporus","Ctropicalis","Cdubliniensis","Calbicans","Sstipitis","Ncastellii","Cglabrata")))
```

```{r plot_prot_length, fig.width=8, fig.height=6}
# plot the length distribution of the homologs in each species
ggplot(seqInfo, aes(x = species, y = length)) + 
  geom_point() + 
  stat_summary(fun.data = mean_cl_boot, geom = "crossbar", col = "red", width = 0.5) + 
  coord_flip() +
  labs(title = "Distribution of XP_028889033 homologs' protein length",
       subtitle = "thick red line indicates group mean and the box the 95% confidence interval by bootstrapping"
       )
```
_Discussion_

- Homologs in the MDR group tend to be longer.

### Adhesin predictors
Now let's import the fungalRV and FaaPred results.
```{r adhesin_prediction}
frv.th = 0.511 # recommended FungalRV score threshold
frv <- read_tsv("raw-output/fungalRV_result.txt", skip = 5, col_names = c("name","frv.score"), col_types = "cd") %>% 
  mutate(name = str_sub(name, 2), frv.pred = frv.score > frv.th)
faa <- read_tsv("raw-output/faapred_result.txt", col_names = c("name","faa.score","faa.pred"), col_types = "cdc") %>% 
  mutate(faa.pred = ifelse(faa.pred == "Adhesin", TRUE, FALSE))
if("frv.score" %in% names(seqInfo))
  seqInfo <- select(seqInfo, -frv.score, -frv.pred, -faa.score, -faa.pred)
seqInfo <- seqInfo %>% left_join(frv) %>% left_join(faa)
seqInfo %>% 
  group_by(species) %>% 
  summarize(n = n(), mean.frv = mean(frv.score), mean.faa = mean(faa.score), fungalRV = sum(frv.score > 0.511), faapred = sum(faa.pred), both = sum(frv.score > 0.55 & faa.pred))
```



### GPI-anchor prediction

```{r gpi}
# GPI-SOM
tmp <- read_delim("raw-output/gpi-anchored-list.txt", delim = " ", col_names = FALSE) %>% pull(X1)
gpi.som <- str_sub(tmp, 2)
if("gpi.som" %in% names(seqInfo))
  seqInfo <- select(seqInfo, -gpi.som)
seqInfo <- mutate(seqInfo, gpi.som = ifelse(name %in% gpi.som, TRUE, FALSE))
# FungalRV pattern matching (this is based on the FungalRV pattern for fungal GPI-anchor sequences)
# this is not very useful as all 110 sequences have >= 1 hit.

# Pred-GPI http://gpcr.biocomp.unibo.it/predgpi
tmp <- read_delim("raw-output/predgpi_result.txt", delim = "|", col_names = c("name","fp","omega"))
pred.gpi <- tmp %>% 
  mutate(name = str_sub(name,2,-2), # remove > and the trailing space
         fp = as.numeric(str_sub(fp, 9, -2)), # extract the numeric part
         pred.gpi = fp <= 0.05,
         omega = str_sub(omega, 7)
         )

if("pred.gpi" %in% names(seqInfo))
  seqInfo <- select(seqInfo, -pred.gpi)
seqInfo <- left_join(seqInfo, pred.gpi %>% select(name, pred.gpi), by = c("name"="name"))

seqInfo %>% 
  group_by(species) %>% 
  summarize(Total = n(), GPI_SOM = sum(gpi.som), GPI_Pred = sum(pred.gpi), Both = sum(gpi.som & pred.gpi))
```

