---
title: "B8441 Hil family PF11765 domain evolution"
author: Bin He
date: "2022-07-07  Updated `r format(Sys.time(), '%d %B, %Y')`"
output: 
  html_notebook:
    toc: true
    toc_float: true
    code_folding: hide
---

```{r}
suppressPackageStartupMessages(require(tidyverse))
suppressPackageStartupMessages(require(cowplot))
suppressPackageStartupMessages(require(ggtext))
suppressPackageStartupMessages(require(ggtree))
suppressPackageStartupMessages(require(treeio))
```

# Goal

- Infer forces of natural selection acting on the PF11765 (NTD) among the Hil homologs in _C. auris_, B8441 strain.
- Differences between this and the previous [MDR-PF11765-evolution](./MDR-PF11765-evolution.Rmd) analysis include
    1. Inclusion of an outgroup homolog from _M. bicuspidata_ to root the domain tree.
    1. Apply PAML on part of the domain alignment that is inferred to be non-recombining by [GARD](https://datamonkey.org)

# Approach

Use PAML to estimate the branch length and omega values for the different branches in the PF11765 tree. If initial results suggest different rates / constraints among the branches, move on to perform formal hypotheses testing.

To avoid making mistakes in manually writing the control files and copying values from the result files to a spreadsheet, I learned to use Biopython's PAML interface to set up a codeml object and write the settings to a control file as well as read the results.

# Data

See `../output/gard/README.md` and other notes in this folder for details on how I generated the partitioned alignment and the tree for each of them.

# Partition 1-414
## One ratio vs free ratio
Read in the tree files with PAML results
```{r}
p1.freeR.cf2 <- read.codeml_mlc("../output/paml/B8441-OG-part/p1-414/branch-freer-cf2/mlc") %>% root(outgroup = "OG")
p1.oneR.cf2 <- read.codeml_mlc("../output/paml/B8441-OG-part/p1-414/branch-1r-cf2/mlc") %>% root(outgroup = "OG")
p1.freeR.cf1 <- read.codeml_mlc("../output/paml/B8441-OG-part/p1-414/branch-freer-cf1/mlc") %>% root(outgroup = "OG")
p1.oneR.cf1 <- read.codeml_mlc("../output/paml/B8441-OG-part/p1-414/branch-1r-cf1/mlc") %>% root(outgroup = "OG")
```
Plot the tree with omega ratios labeled for branches with high dN/dS
```{r}
p.cf2 <- ggtree(p1.freeR.cf2, aes(color = dN_vs_dS), size = 1, branch.length = 'none') + 
  #geom_rootedge(rootedge = 0.1) +
  geom_nodelab(hjust = -0.2, size = 3, color = "black") +
  geom_tiplab(size = 6, hjust = -0.5, color = 1) + 
  xlim(0, 5.6) +
  geom_text(data = td_filter(dN_vs_dS > 1),
            aes(x = branch, label = paste0("ω = ", round(dN_vs_dS, 0))), 
            vjust = -.5, size = 4) + 
  geom_text(data = td_filter(dN_vs_dS > 1),
            aes(x = branch, label = paste0("dN = ", round(dN, 2), "\n", "dS = ", round(dS, 2))), 
            vjust = 1.2, lineheight = 0.8, size = 4) + 
  geom_richtext(data = td_filter(node == 12), aes(x = branch, label = "ω1"), 
                fill = NA, label.color = NA, vjust = -0.5, size = 6, color = "red") +
  geom_richtext(data = td_filter(node == 11), aes(x = branch, label = "ω2"), 
                fill = NA, label.color = NA, vjust = -0.5, size = 6, color = "red") +
  geom_richtext(data = td_filter(node == 16), aes(x = branch, label = "ω3"), 
                fill = NA, label.color = NA, vjust = -0.5, size = 6, color = "red") +
  scale_color_continuous(name='dN/dS', limits=c(0, 1.5), 
                         oob=scales::squish, low='darkgreen', high="darkviolet") +
  theme(legend.position = c(0.1,0.8), legend.text = element_text(size = 12), 
        legend.title = element_text(size = 14))
#ggsave("../output/figure/20220122-MDR-PF11765-freeR-anno-tree.png", width = 7, height = 4.5)
```

```{r}
p.cf1 <- ggtree(p1.freeR.cf1, aes(color = dN_vs_dS), size = 1, branch.length = 'none') + 
  #geom_rootedge(rootedge = 0.1) +
  geom_nodelab(hjust = -0.2, size = 3, color = "black") +
  geom_tiplab(size = 6, hjust = -0.5, color = 1) + 
  xlim(0, 5.6) +
  geom_text(data = td_filter(dN_vs_dS > 1),
            aes(x = branch, label = paste0("ω = ", round(dN_vs_dS, 0))), 
            vjust = -.5, size = 4) + 
  geom_text(data = td_filter(dN_vs_dS > 1),
            aes(x = branch, label = paste0("dN = ", round(dN, 2), "\n", "dS = ", round(dS, 2))), 
            vjust = 1.2, lineheight = 0.8, size = 4) + 
  geom_richtext(data = td_filter(node == 12), aes(x = branch, label = "ω1"), 
                fill = NA, label.color = NA, vjust = -0.5, size = 6, color = "red") +
  geom_richtext(data = td_filter(node == 13), aes(x = branch, label = "ω2"), 
                fill = NA, label.color = NA, vjust = -0.5, size = 6, color = "red") +
  geom_richtext(data = td_filter(node == 16), aes(x = branch, label = "ω3"), 
                fill = NA, label.color = NA, vjust = -0.5, size = 6, color = "red") +
  geom_richtext(data = td_filter(node == 9), aes(x = branch, label = "ω4"), 
                fill = NA, label.color = NA, vjust = -0.5, size = 6, color = "red") +
  scale_color_continuous(name='dN/dS', limits=c(0, 1.5), 
                         oob=scales::squish, low='darkgreen', high="darkviolet") +
  theme(legend.position = c(0.1,0.8), legend.text = element_text(size = 12), 
        legend.title = element_text(size = 14))
#ggsave("../output/figure/20220122-MDR-PF11765-freeR-anno-tree.png", width = 7, height = 4.5)
```
```{r fig.width = 10, fig.height = 5}
plot_grid(p.cf1 + labs(title = "CodonFreq = 1"), p.cf2 + labs(title = "CodonFreq = 2"), nrow = 1)
ggsave(paste0("../output/figure/", gsub("-", "", Sys.Date()),"-B8441-OG-p1-414-branch-freer-tree.png"), width = 10, height = 5)
```

> In my previous analysis where I found that how I specify the codonfreq parameter has a profound effect on the branch lengths estimates. In particular, `CodonFreq=2` (F3x4) results in very large dS estimates and identified different branches as having elevated dN/dS than `CodonFreq=0/1` did. Here focusing on the 1-414 fragment, I found the inconsistencies between `CodonFreq=2` and `CodonFreq=1` are less significant. A few branches were identified as having large ω by both analyses:
> 
> - 12..16 (ancestral to Hil6 and Hil8)
> - 11..12 (ancestral to Hil1, 2, 3, 4, 6 and 8)

Note that `CodonFreq=2` results in a highly non-ultrametric tree
```{r}
ggtree(p1.freeR.cf2, aes(color = dS)) + 
  scale_color_gradient(trans = "sqrt", breaks = c(0, 1, 10, 30, 60)) +
  geom_tiplab(color = 1) + geom_nodelab(color = 1, size = 3, hjust = -.2)
```

This is not the case for `CodonFreq=1`
```{r}
ggtree(p1.freeR.cf1, aes(color = dS)) + 
  scale_color_viridis_c(option = "D", trans = "sqrt") +
  geom_tiplab(color = 1) + geom_nodelab(color = 1, size = 3, hjust = -.2)
```

# Partition 697-981
## One ratio vs free ratio
Read in the tree files with PAML results
```{r}
p2.freeR.cf2 <- read.codeml_mlc("../output/paml/B8441-OG-part/p697-981/branch-freer-cf2/mlc") %>% root(outgroup = "OG")
p2.oneR.cf2 <- read.codeml_mlc("../output/paml/B8441-OG-part/p697-981/branch-1r-cf2/mlc") %>% root(outgroup = "OG")
p2.freeR.cf1 <- read.codeml_mlc("../output/paml/B8441-OG-part/p697-981/branch-freer-cf1/mlc") %>% root(outgroup = "OG")
p2.oneR.cf1 <- read.codeml_mlc("../output/paml/B8441-OG-part/p697-981/branch-1r-cf1/mlc") %>% root(outgroup = "OG")
```
Plot the tree with omega ratios labeled for branches with high dN/dS
```{r}
p.cf2 <- ggtree(p2.freeR.cf2, aes(color = dN_vs_dS), size = 1, branch.length = 'none') + 
  #geom_rootedge(rootedge = 0.1) +
  geom_nodelab(hjust = -0.2, size = 3, color = "black") +
  geom_tiplab(size = 6, hjust = -0.5, color = 1) + 
  xlim(0, 6) +
  geom_text(data = td_filter(dN_vs_dS > 1),
            aes(x = branch, label = paste0("ω = ", round(dN_vs_dS, 0))), 
            vjust = -.5, size = 3) + 
  geom_text(data = td_filter(dN_vs_dS > 1),
            aes(x = branch, label = paste0("dN = ", round(dN, 2), "\n", "dS = ", round(dS, 2))), 
            vjust = 1.2, lineheight = 0.8, size = 3) + 
  #geom_richtext(data = td_filter(node == 12), aes(x = branch, label = "ω1"), 
  #              fill = NA, label.color = NA, vjust = -0.5, size = 6, color = "red") +
  #geom_richtext(data = td_filter(node == 11), aes(x = branch, label = "ω2"), 
  #              fill = NA, label.color = NA, vjust = -0.5, size = 6, color = "red") +
  #geom_richtext(data = td_filter(node == 16), aes(x = branch, label = "ω3"), 
  #              fill = NA, label.color = NA, vjust = -0.5, size = 6, color = "red") +
  scale_color_continuous(name='dN/dS', limits=c(0, 1.5), 
                         oob=scales::squish, low='darkgreen', high="darkviolet") +
  theme(legend.position = c(0.1,0.8), legend.text = element_text(size = 12), 
        legend.title = element_text(size = 14))
p.cf2
```

```{r}
p.cf1 <- ggtree(p2.freeR.cf1, aes(color = dN_vs_dS), size = 1, branch.length = 'none') + 
  #geom_rootedge(rootedge = 0.1) +
  geom_nodelab(hjust = -0.2, size = 3, color = "black") +
  geom_tiplab(size = 6, hjust = -0.5, color = 1) + 
  xlim(0, 6) +
  geom_text(data = td_filter(dN_vs_dS > 1),
            aes(x = branch, label = paste0("ω = ", round(dN_vs_dS, 0))), 
            vjust = -.5, size = 3) + 
  geom_text(data = td_filter(dN_vs_dS > 1),
            aes(x = branch, label = paste0("dN = ", round(dN, 2), "\n", "dS = ", round(dS, 2))), 
            vjust = 1.2, lineheight = 0.8, size = 3) + 
  #geom_richtext(data = td_filter(node == 12), aes(x = branch, label = "ω1"), 
  #              fill = NA, label.color = NA, vjust = -0.5, size = 6, color = "red") +
  #geom_richtext(data = td_filter(node == 13), aes(x = branch, label = "ω2"), 
  #              fill = NA, label.color = NA, vjust = -0.5, size = 6, color = "red") +
  #geom_richtext(data = td_filter(node == 16), aes(x = branch, label = "ω3"), 
  #              fill = NA, label.color = NA, vjust = -0.5, size = 6, color = "red") +
  #geom_richtext(data = td_filter(node == 9), aes(x = branch, label = "ω4"), 
  #              fill = NA, label.color = NA, vjust = -0.5, size = 6, color = "red") +
  scale_color_continuous(name='dN/dS', limits=c(0, 1.5), 
                         oob=scales::squish, low='darkgreen', high="darkviolet") +
  theme(legend.position = c(0.1,0.8), legend.text = element_text(size = 12), 
        legend.title = element_text(size = 14))
p.cf1
```

```{r fig.width = 10, fig.height = 5}
plot_grid(p.cf1 + labs(title = "CodonFreq = 1"), p.cf2 + labs(title = "CodonFreq = 2"), nrow = 1)
ggsave(paste0("../output/figure/", gsub("-", "", Sys.Date()),"-B8441-OG-p697-981-branch-freer-tree.png"), width = 10, height = 5)
```
