---
title: "B8441 Hil family PF11765 domain evolution"
author: Bin He
date: "2022-07-07  Updated `r format(Sys.time(), '%d %B, %Y')`"
output: 
  html_notebook:
    toc: true
    toc_float: true
    code_folding: hide
---

```{r}
suppressPackageStartupMessages(require(tidyverse))
suppressPackageStartupMessages(require(cowplot))
suppressPackageStartupMessages(require(ggtree))
suppressPackageStartupMessages(require(treeio))
```

# Goal

- Infer forces of natural selection acting on the PF11765 (NTD) among the Hil homologs in _C. auris_, B8441 strain.
- Differences between this and the previous [MDR-PF11765-evolution](./MDR-PF11765-evolution.Rmd) analysis include
    1. Inclusion of an outgroup homolog from _M. bicuspidata_ to root the domain tree.
    1. Apply PAML on part of the domain alignment that is inferred to be non-recombining by [GARD](https://datamonkey.org)

# Approach

Use PAML to estimate the branch length and omega values for the different branches in the PF11765 tree. If initial results suggest different rates / constraints among the branches, move on to perform formal hypotheses testing.

To avoid making mistakes in manually writing the control files and copying values from the result files to a spreadsheet, I learned to use Biopython's PAML interface to set up a codeml object and write the settings to a control file as well as read the results.

# Data

See `../output/gard/README.md` and other notes in this folder for details on how I generated the partitioned alignment and the tree for each of them.

# Partition 1-414
## One ratio vs free ratio
Read in the tree files with PAML results
```{r}
p1.freeR.cf2 <- read.codeml_mlc("../output/paml/B8441-OG-part/p1-414/branch-freer-cf2/mlc") %>% root(outgroup = "OG")
p1.oneR.cf2 <- read.codeml_mlc("../output/paml/B8441-OG-part/p1-414/branch-1r-cf2/mlc") %>% root(outgroup = "OG")
p1.freeR.cf1 <- read.codeml_mlc("../output/paml/B8441-OG-part/p1-414/branch-freer-cf1/mlc") %>% root(outgroup = "OG")
p1.oneR.cf1 <- read.codeml_mlc("../output/paml/B8441-OG-part/p1-414/branch-1r-cf1/mlc") %>% root(outgroup = "OG")
```
Plot the tree with omega ratios labeled for branches with high dN/dS
```{r}
p.cf2 <- ggtree(p1.freeR.cf2, aes(color = dN_vs_dS), size = 1, branch.length = 'none') + 
  #geom_rootedge(rootedge = 0.1) +
  geom_nodelab(hjust = -0.2, size = 3, color = "black") +
  geom_tiplab(size = 6, hjust = -0.5, color = 1) + 
  xlim(0, 5.6) +
  geom_text(data = td_filter(dN_vs_dS > 0.5),
            aes(x = branch, label = paste0("ω = ", round(dN_vs_dS, 0))), 
            vjust = -.5, size = 4) + 
  geom_text(data = td_filter(dN_vs_dS > 0.5),
            aes(x = branch, label = paste0("dN = ", round(dN, 2), "\n", "dS = ", round(dS, 2))), 
            vjust = 1.2, lineheight = 0.8, size = 4) + 
  #geom_text(data = td_filter(node == 15), aes(x = branch, label = "ω1"), 
  #          vjust = 1.5, size = 6, color = "black") +
  #geom_text(data = td_filter(node == 11), aes(x = branch, label = "ω2"), 
  #          vjust = 1.5, size = 6, color = "black") +
  #geom_text(data = td_filter(node %in% c(4, 14)), aes(x = branch, label = "ω3"), 
  #          vjust = 1.5, size = 6, color = "black") +
  scale_color_continuous(name='dN/dS', limits=c(0, 1.5), 
                         oob=scales::squish, low='darkgreen', high="darkviolet") +
  theme(legend.position = c(0.1,0.8), legend.text = element_text(size = 12), 
        legend.title = element_text(size = 14))
#ggsave("../output/figure/20220122-MDR-PF11765-freeR-anno-tree.png", width = 7, height = 4.5)
```

```{r}
p.cf1 <- ggtree(p1.freeR.cf1, aes(color = dN_vs_dS), size = 1, branch.length = 'none') + 
  #geom_rootedge(rootedge = 0.1) +
  geom_nodelab(hjust = -0.2, size = 3, color = "black") +
  geom_tiplab(size = 6, hjust = -0.5, color = 1) + 
  xlim(0, 5.6) +
  geom_text(data = td_filter(dN_vs_dS > 0.5),
            aes(x = branch, label = paste0("ω = ", round(dN_vs_dS, 0))), 
            vjust = -.5, size = 4) + 
  geom_text(data = td_filter(dN_vs_dS > 0.5),
            aes(x = branch, label = paste0("dN = ", round(dN, 2), "\n", "dS = ", round(dS, 2))), 
            vjust = 1.2, lineheight = 0.8, size = 4) + 
  #geom_text(data = td_filter(node == 15), aes(x = branch, label = "ω1"), 
  #          vjust = 1.5, size = 6, color = "black") +
  #geom_text(data = td_filter(node == 11), aes(x = branch, label = "ω2"), 
  #          vjust = 1.5, size = 6, color = "black") +
  #geom_text(data = td_filter(node %in% c(4, 14)), aes(x = branch, label = "ω3"), 
  #          vjust = 1.5, size = 6, color = "black") +
  scale_color_continuous(name='dN/dS', limits=c(0, 1.5), 
                         oob=scales::squish, low='darkgreen', high="darkviolet") +
  theme(legend.position = c(0.1,0.8), legend.text = element_text(size = 12), 
        legend.title = element_text(size = 14))
#ggsave("../output/figure/20220122-MDR-PF11765-freeR-anno-tree.png", width = 7, height = 4.5)
```
```{r}
plot_grid(as_)
```

## _C. auris_ free ratio
I found how I specify the codonfreq parameter has a profound effect on the branch lengths estimates, and to some extent on the dN/dS estimates as well, where `CodonFreq=2` (F3x4) results in very large dS estimates and identified different branches as having elevated dN/dS than `CodonFreq=0/1` did.

### `CodonFreq = 0`
```{r}
caur.freeR.cf0 <- read.codeml_mlc("../output/paml/B8441-Hil1-8-PF11765/codonfreq0-freeR/mlc")

ggtree(caur.freeR.cf0, aes(color = dN_vs_dS), size = 1, branch.length = 'none') + 
  #geom_rootedge(rootedge = 0.1) +
  geom_nodelab(hjust = -0.2, size = 3, color = "black") +
  geom_tiplab(aes(label = str_extract(label, "Hil[1-8]")), size = 6, hjust = -0.5, color = 1) + 
  xlim(0, 5.6) +
  geom_text(data = td_filter(dN_vs_dS > 0.5),
            aes(x = branch, label = paste(round(dN, 2), round(dS, 2), sep = "/")), 
            vjust = -.5, size = 4) + 
  geom_text(data = td_filter(node == 15), aes(x = branch, label = "ω1"), 
            vjust = 1.5, size = 6, color = "black") +
  geom_text(data = td_filter(node == 11), aes(x = branch, label = "ω2"), 
            vjust = 1.5, size = 6, color = "black") +
  geom_text(data = td_filter(node %in% c(4, 14)), aes(x = branch, label = "ω3"), 
            vjust = 1.5, size = 6, color = "black") +
  scale_color_continuous(name='dN/dS', limits=c(0, 1.5), 
                         oob=scales::squish, low='darkgreen', high="darkviolet") +
  theme(legend.position = c(0.1,0.8), legend.text = element_text(size = 12), 
        legend.title = element_text(size = 14))
ggsave("../output/figure/20220122-caur-PF11765-Feual-freeR-anno-tree.png", width = 5, height = 5)
```

### `CodonFreq = 1`
```{r}
caur.freeR.cf1 <- read.codeml_mlc("../output/paml/B8441-Hil1-8-PF11765/codonfreq1-freeR/mlc")
ggtree(caur.freeR.cf1, aes(color = dN_vs_dS), size = 1, branch.length = 'none') + 
  #geom_rootedge(rootedge = 0.1) +
  geom_nodelab(hjust = -0.2, size = 3, color = "black") +
  geom_tiplab(aes(label = str_extract(label, "Hil[1-8]")), size = 6, hjust = -0.5, color = 1) + 
  xlim(0, 5.6) +
  geom_text(data = td_filter(dN_vs_dS > 0.5),
            aes(x = branch, label = paste(round(dN, 2), round(dS, 2), sep = "/")), 
            vjust = -.5, size = 4) + 
  geom_text(data = td_filter(node == 15), aes(x = branch, label = "ω1"), 
            vjust = 1.5, size = 6, color = "black") +
  geom_text(data = td_filter(node == 11), aes(x = branch, label = "ω2"), 
            vjust = 1.5, size = 6, color = "black") +
  geom_text(data = td_filter(node %in% c(4, 14)), aes(x = branch, label = "ω3"), 
            vjust = 1.5, size = 6, color = "black") +
  scale_color_continuous(name='dN/dS', limits=c(0, 1.5), 
                         oob=scales::squish, low='darkgreen', high="darkviolet") +
  theme(legend.position = c(0.1,0.8), legend.text = element_text(size = 12), 
        legend.title = element_text(size = 14))
ggsave("../output/figure/20220122-caur-PF11765-F1x4-freeR-anno-tree.png", width = 5, height = 5)
```

### `CodonFreq = 2`
Plotted as cladogram to avoid compressing some of the lineages. Branch lengths are printed at the bottom of the branches.
```{r}
caur.freeR.cf2 <- read.codeml_mlc("../output/paml/B8441-Hil1-8-PF11765/codonfreq2-freeR/mlc")
# several internal branches have 0 inferred syn or nonsyn changes, and the dN/dS ratio estimates doesn't make much sense
ggtree(caur.freeR.cf2, aes(color = dN_vs_dS), branch.length = 'none') + 
  #geom_rootedge(rootedge = 0.1) +
  geom_nodelab(hjust = -0.2, size = 3, color = "black") +
  geom_tiplab(aes(label = str_extract(label, "Hil[1-8]")), size = 6, hjust = -0.5, color = 1) + 
  xlim(0, 5.6) +
  geom_text(aes(x = branch, label = round(t, 3)), vjust = 1.5, color = "black", size = 3) + 
  geom_text(aes(x = branch, label = paste(round(dN, 2), round(dS, 2), sep = " / ")), 
            vjust = -.5, size = 4) + 
  scale_color_continuous(name='dN/dS', limits=c(0, 1.5), 
                         oob=scales::squish, low='darkgreen', high="darkviolet") +
  theme(legend.position = c(0.1,0.8), legend.text = element_text(size = 12), 
        legend.title = element_text(size = 14))
  # geom_nodelab(hjust = -0.2, size = 2, color = "black") +
  # geom_tiplab(as_ylab = TRUE) +
  # geom_text(aes(x = branch, label = paste(round(dN, 2), round(dS, 2), sep = " / ")), vjust = -.5) + 
  # scale_color_continuous(name='dN/dS', limits=c(0, 1.5),
  #                        oob=scales::squish,
  #                        low='darkgreen', high='red')
ggsave("../output/figure/20220130-caur-PF11765-F3x4-freeR-anno-tree.png", width = 5, height = 4)
```
