---
title: "Analyze XP_028889033 family evolution and adhesin properties"
author: "Bin He"
date: "11/01/2020"
output: 
  html_notebook:
    toc: yes
    toc_float: yes
    toc_depth: 5
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r load_libraries, echo = FALSE}
#if (!requireNamespace("BiocManager", quietly = TRUE))
#    install.packages("BiocManager")
#if (!requireNamespace("treeio", quietly = TRUE))
#    BiocManager::install("treeio")
#if (!requireNamespace("ggtree", quietly = TRUE))
#    BiocManager::install("ggtree")
library(tidyverse)
library(cowplot)
```

## Goal

Analyze the adhesin properties of the XP_028889033 homologs (putative adhesin in _C. auris_)
This is version 3 of the analysis, using the updated homologs on 2020-10-31

## Build datasets
First get the basic information about the 100 sequences in this study. I decide to write a simple Python script to extract such info.
```{bash}
# edit the PYTHON path below to match your local system
~/sw/miniconda3/bin/python extract_seq_info.py
```

Load in the sequence information.
```{r load_seq_info}
seqInfo <- read_tsv("raw-output/XP_028889033_homologs.tsv", comment = "#", col_types = "ccci") %>% 
  mutate(species_id = factor(species, levels = c("Cduobushaemulonis","Cpseudohaemulonis","Chaemuloni","Cauris","Clusitaniae","Dhansenii","Cparapsilosis","Lelongisporus","Ctropicalis","Cdubliniensis","Calbicans","Sstipitis","Klactis","Ncastellii","Cglabrata","Nbracarensis","Ndelphensis","Nnivariensis")), species = NULL)
chrLoc <- read_tsv("raw-output/XP_028889033_homologs_chr_loc.tsv") %>% 
  mutate(id = ifelse(is.na(accession), gene_id, accession), accession = NULL, species_id = factor(species_id, levels = c("Cduobushaemulonis","Cpseudohaemulonis","Chaemuloni","Cauris","Clusitaniae","Dhansenii","Cparapsilosis","Lelongisporus","Ctropicalis","Cdubliniensis","Calbicans","Sstipitis","Klactis","Ncastellii","Cglabrata","Nbracarensis","Ndelphensis","Nnivariensis")))
seqInfo <- seqInfo %>%  left_join(chrLoc) %>% 
  mutate(assemblystatus = factor(assemblystatus, levels = c("Complete Genome","Chromosome","Contig","Scaffold"), 
                                 labels = c("Chromosome", "Chromosome", "Partial", "Partial")),
         species_gr = factor(species_id, labels = c(rep("MDR",4),rep("Clavispora",2),rep("albicans",5), "Sstipitis", "Klactis", rep("Nakaseomycetes",5)))) %>% 
  select(name, id, source, starts_with("gene"), starts_with("species"), strain, starts_with("assembly"),
         length, n_seqs, starts_with("chr"), relLoc)
```

## Homologous protein length
```{r plot_prot_length, fig.width=8, fig.height=5}
# plot the length distribution of the homologs in each species
ggplot(seqInfo, aes(x = species, y = length)) + 
  geom_boxplot() + geom_hline(yintercept = 500, linetype = 2) +
  coord_flip() +
  labs(title = "Distribution of XP_028889033 homologs' protein length")
```
_Discussion_

- The one sequence below 500 a.a. is from _N. delphensis_, which is labeled as a partial CDS.
- Majority of the proteins in the list are 500-2000 a.a., with a few exceptionally long

## Chromosomal locations
While I was performing BLAST on FungiDB to identify homologs of our protein, I noticed that many of the hits appear to be at the beginning and end of the chromosomes. To see if there is a systematic trend in the chromosomal locations, I collected this information for all 99 homologs in the list. Note however, not all genomes are assembled to the chromosomal level, and as a result, the locations for the proteins in those species/strains would be relative to a contig or scaffold. My rationale is that if the contig or scaffold is close to choromosomal length (although that varies by at least an order of magnitude), I could at least look at them separately.

```{r, fig.width = 8, fig.height = 6}
p <- seqInfo %>% 
  mutate(chr = paste(species_id, chraccver, sep = "_")) %>% 
  ggplot(aes(x = chr)) + 
  geom_segment(aes(xend = chr, y = 1, yend = chrL, color = assemblystatus), size = 2) +
  geom_segment(aes(xend = chr, y = chrstart, yend = chrstart+5000), color = "red", size = 2)
p + coord_flip() + theme_classic() + scale_color_manual(values = c("grey50", "grey")) +
  theme(axis.text.y = element_text(size = 5),
        axis.line.y = element_blank(), axis.ticks.y = element_blank(),
        axis.line.x = element_blank(), axis.ticks.x = element_blank(),
        legend.position = c(0.85,0.85),
        panel.background = element_rect(fill = alpha("lightblue",0.5))) +
  labs(y = "Position on chromosome or scaffold", x = "Chromosomes / Scaffolds", color = "Assembly Status") + 
  ggtitle("Chromosomal locations of XP_028889033 homologs")
ggsave("img/20201102-homologs-chr-loc.png", bg = "transparent", width = 7, height = 7)
```

```{r, fig.width = 8, fig.height = 4}
p <- seqInfo %>% filter(!(chrL < 1e6 & assemblystatus == "Partial")) %>% 
  mutate(dTip = ifelse(relLoc < 0.5, relLoc, 1-relLoc)) %>% 
  ggplot(aes(x = dTip)) + facet_wrap(~assemblystatus) +
  xlab("Distance from Chromosome or Scaffold Ends") + labs(fill = "Species group") + 
  scale_x_continuous(breaks = seq(0,0.5,by = 0.1))
p + geom_histogram(binwidth = 0.05)
ggsave("img/20201102-distribution-on-chromosome-or-scaffolds-histogram-grey.png", width = 5, height = 3)
p + geom_histogram(aes(fill = species_gr), binwidth = 0.05) + scale_fill_brewer(palette = "Paired")
ggsave("img/20201102-distribution-on-chromosome-or-scaffolds-histogram.png", width = 6, height = 3)
```

**Discussion**

- Visually there appear to be some enrichment towards the chromosome ends overall.
- Breakdown by species revealed stronger signals in the albicans clade and the MDR clade.
- To properly test for enrichment in the chromosomal ends would require documenting the gene density on the affected chromosomes. If the different chromosomes have dramatically different gene distributions, proper tests may not be straightforward to construct. Instead, a bootstrapping approach could be easier to conceive and apply.

## Adhesin predictors
Now let's import the fungalRV and FaaPred results.
```{r adhesin_prediction}
frv.th = 0.511 # recommended FungalRV score threshold
frv <- read_tsv("raw-output/fungalRV_result.txt", skip = 3, col_names = c("name","frv.score"), col_types = "cd") %>% 
  mutate(name = str_sub(name, 2), frv.pred = frv.score > frv.th)
faa <- read_tsv("raw-output/faapred_result.txt", col_names = c("name","faa.score","faa.pred"), col_types = "cdc") %>% 
  mutate(faa.pred = ifelse(faa.pred == "Adhesin", TRUE, FALSE))
if("frv.score" %in% names(seqInfo))
  seqInfo <- select(seqInfo, -frv.score, -frv.pred, -faa.score, -faa.pred)
seqInfo <- seqInfo %>% left_join(frv) %>% left_join(faa)
seqInfo %>% 
  group_by(species) %>% 
  summarize(n = n(), mean.frv = mean(frv.score), mean.faa = mean(faa.score, na.rm = T), 
            fungalRV = sum(frv.score > 0.511), faapred = sum(faa.pred, na.rm = T), 
            both = sum(frv.score > 0.511 & faa.pred))
```


## GPI-anchor prediction
GPI-anchored proteins are characterized by an N-terminal signal peptide, which would direct the protein to the secretary pathway, and a C-terminal GPI-anchor peptide, which would be cleaved and replaced by the GPI-anchor, allowing the protein to be tethered to the cell wall. For signal peptide, I used SignalP server. Its latest version is 5.0. But I also ran the sequences through their 4.1 version, with two settings. The results of the latter two are almost identical, except for one sequence "XP_024711350.1", which is only included in the sensitive version, and has a probability lower than 0.5.

```{r signalP, fig.width=5, fig.height=5}
# Signal peptide
gff.names <- c("id", "source", "name", "start", "end", "prob", "na1", "na2", "na3")
signalp5 <- read_tsv("raw-output/signalp_5.0.gff", comment = "#", col_names = gff.names, col_types = "ccciidccc")

if("signalp" %in% names(seqInfo))
  seqInfo <- select(seqInfo, -signalp)

seqInfo <- left_join(seqInfo, select(signalp5, name = id, prob), by = c("name" = "name")) %>% 
  mutate(signalp = !is.na(prob)) %>% select(-prob)
```

For GPI-anchor prediction, I used the [PredGPI server](http://gpcr.biocomp.unibo.it/predgpi/).
```{r gpi}
tmp <- read_delim("raw-output/predgpi_result.txt", delim = "|", col_names = c("name","fp","omega"))
pred.gpi <- tmp %>% 
  mutate(name = str_sub(name,2,-2), # remove > and the trailing space
         fp = as.numeric(str_sub(fp, 9, -2)), # extract the numeric part
         is.gpi = fp <= 0.01,    # based on the cutoff of the PredGPI server (prob < 99% -> not GPI-anchored)
         omega = str_sub(omega, 8),
         cleaveRes = str_sub(omega, 1, 1),
         cleavePos = as.integer(str_sub(omega, 3))
         ) %>% 
  left_join(select(seqInfo, name, length), by = c("name" = "name"))

# remove the column if it already exists
if("pred.gpi" %in% names(seqInfo))
  seqInfo <- select(seqInfo, -pred.gpi)
seqInfo <- left_join(seqInfo, select(pred.gpi, name, pred.gpi = is.gpi), by = c("name"="name"))

seqInfo %>% 
  group_by(species) %>% 
  summarize(Total = n(), SignalP = sum(signalp), GPI_Pred = sum(pred.gpi), Both = sum(signalp & pred.gpi))
```

## Build feature map for homologs

The goal is to produce a cartoon-like plot for each homolog outlining their main features, such as the locations of the PFam domains (mainly the Hyp_reg_CWP), locations of the signal peptide and GPI-anchor, distribution of TANGO sequences. Note that all these features can be represented as a range with associated metadata. So the first step is to collect the coordinates of the features

```{r}
# GPI-anchor
# use pred.gpi
# Pfam domains
pfam <- read_tsv("raw-output/HMMER-HMMScan-Pfam-hits.tsv", col_types = "ciiiicciiidddiic")
# save feature file for Jalview examination
# pfam %>% filter(grepl("XP_028889033",seq_id)) %>% select(hmm_name, seq_id, envelope_start, envelope_end) %>% mutate(featuretype = "domain") %>% write_tsv("XP_028889033_features.jalview")
# I manually edited the feature file, so I commented out the line above to avoid accidentally 
# overwriting my own edits

# feature set
# structure: id  feature  start  end
feature <- bind_rows(
  seqInfo %>% mutate(type = "entire protein", start = 1) %>% select(id = name, type, start, end = length),
  pfam %>% select(id = seq_id, type = hmm_name, start = envelope_start, end = envelope_end),
  # extend the signal peptide segment by 10 amino acids to make it more visible
  signalp5 %>% mutate(type = "SignalP", end = end + 10) %>% select(id, type, start = start, end),
  # extend the GPI-anchor C-terminus segment by 20 amino acids to make it more visible
  pred.gpi %>% filter(is.gpi) %>% mutate(type = "GPI-anchor", start = cleavePos-10) %>% 
    select(id = name, type, start, end = length)
)
feature$type = ordered(feature$type, levels = c("entire protein", "Hyphal_reg_CWP", "Asp", "Candida_ALS", "Flocculin", "Flocculin_t3", "Hyr1", "PT", "SignalP", "GPI-anchor"))
feature.colors <- c("grey", "#1f78b4", "#b2df8a", "#33a02c", "#fb9a99", "#fdbf6f", "#ff7f00", "#cab2d6", "#e31a1c", "#6a3d9a")
# in order to plot properties of the sequences in an order that is consistent with the sequences' position in the gene tree
genetreeOrder <- scan("reorder_by_gene_tree.txt", what = "character")
feature$id <- ordered(feature$id, levels = rev(genetreeOrder))
```

```{r plot_features, fig.width=10}
p <- ggplot(feature, aes(x = id, y = start)) + 
  geom_segment(aes(xend = id, yend = end, color = type), size = 2)
p + coord_flip() + theme_classic() + scale_color_manual(values = feature.colors) +
  theme(axis.text.y = element_text(size = 5),
        axis.line.y = element_blank(), axis.ticks.y = element_blank(),
        axis.line.x = element_blank(), axis.ticks.x = element_blank(),
        legend.position = c(0.85,0.25),
        panel.background = element_rect(fill = alpha("lightblue",0.5))) +
  ylim(1, 4500) + labs(y = "Position in sequence", x = "Sequences", color = "Domain / Feature") + 
  ggtitle("Domain architecture")
ggsave("img/20201223-homologs-domains-schematics.png", bg = "transparent", width = 7, height = 7)
```

## Serine/Threonine content

S/T sites are potential sites for O-glycosylation, which could increase the rididity of the stalk of the protein and allow the N-terminal domain to protrude out of the cell wall facing the exterior. More evidence for the importance of O-glycosylation in a serine/threonine-rich domain can be found [here](https://ec.asm.org/content/10/10/1317.long).

To determine the S/T frequency in the XP_028889033 homologs, I ran the program `freak` from the EMBOSS suite with the parameters of 100 aa sliding window and a step size of 10 aa. After reformating the output, the rest of the analysis is accomplished below.

```{r S_T_freq}
# load data
ST.freq <- read_tsv("raw-output/ST_freq_100_10_freak.out.gz", col_types = "cid")
S.freq <- read_tsv("raw-output/S_freq_100_10_freak.out.gz", col_types = "cid")
T.freq <- read_tsv("raw-output/T_freq_100_10_freak.out.gz", col_types = "cid")
# convert sequence name column to an ordered list sorted based on the gene tree sequence
ST.freq <- ST.freq %>%  mutate(id = ordered(id, levels = rev(genetreeOrder))) # this will produce the desired order
S.freq <- S.freq %>%  mutate(id = ordered(id, levels = rev(genetreeOrder))) # this will produce the desired order
T.freq <- T.freq %>%  mutate(id = ordered(id, levels = rev(genetreeOrder))) # this will produce the desired order
```

```{r plot_ST_freq, fig.width=10}
ggplot(ST.freq, aes(x = id, y = pos)) +  geom_tile(aes(fill = freq)) +
  coord_flip() + theme_classic() + scale_fill_distiller(palette = "RdGy", limits = c(0, 0.8), oob = scales::squish, breaks = seq(0, 0.6, by = 0.2)) +
  theme(axis.text.y = element_text(size = 5),
        axis.line.y = element_blank(), axis.ticks.y = element_blank(),
        axis.line.x = element_blank(), axis.ticks.x = element_blank(),
        legend.position = c(0.85,0.25),
        panel.background = element_rect(fill = alpha("lightblue",0.5))) +
  ylim(1, 4500) + labs(y = "Position in sequence", x = "Sequences", color = "Frequency") + 
  ggtitle("Serine/Threonine frequency in 100 aa sliding windows")
ggsave("img/20201223-homologs-ST-freq-100aa-window.png", bg = "transparent", width = 7, height = 7)
```
```{r, fig.width=10}
ST.comb <- bind_rows("S/T" = ST.freq, S = S.freq, T = T.freq, .id = "Var") %>% 
  mutate(Var = ordered(Var, levels = c("S/T","S","T")))
ggplot(ST.comb, aes(x = id, y = pos)) + geom_tile(aes(fill = freq)) + 
  facet_wrap(~Var, scales = "fixed") + theme_minimal() +
  coord_flip() + scale_fill_distiller(palette = "RdGy", limits = c(-0.05, 0.55), oob = scales::squish,
                                      breaks = seq(0,0.5,by=0.1)) +
  #scale_fill_gradient2(low = "#1a1a1a", high = "#b30000", midpoint = 0.2) +
  theme(axis.text.y = element_blank(),
        axis.line.y = element_blank(), axis.ticks.y = element_blank(),
        axis.line.x = element_blank(), axis.ticks.x = element_blank(),
        legend.position = c(0.95,0.24),
        panel.background = element_rect(fill = alpha("lightblue",0.5))) +
  ylim(1, 4500) + ylab("Position in sequence") + xlab("Sequences") + 
  ggtitle("Ser/Thr frequency in 100 aa windows (frequency > 0.55 replaced by 0.55)")
ggsave("img/20201223-ST-freq-composite.png", width = 8, height = 4.5)
```


```{r, eval=FALSE, include=FALSE}
suppressPackageStartupMessages(library(ggtree))
geneTree <- read.tree("XP_028889033_homologs_gene_tree.nwk")
```

## TANGO prediction of $\beta$-aggregation prone sequences

The amyloid-like $\beta$-aggregation prone sequences have the ability to mediate self-aggregation, which boosts the local concentration of the adhesin molecules on the cell-surface. Similar to the S/T frequency above, we would like to use the output from the prediction algorithm, TANGO, to visulize the distribution of such sequence motifs along the length of the XP_028889033 homolog sequences.

TANGO predictions and the subsequent extraction of $\beta$-aggregation prone sequences were documented in separate Python script, README files and the `tango.Rmd`.
```{r tango_sequences, fig.width=10}
# here we only look at the extracted tango sequences
tango <- read_tsv("tango_summary_table.tsv.gz", col_types = "cciiiddicc")
# reorder the sequences for plotting
tango$name <- ordered(tango$name, levels = rev(genetreeOrder))
# plot
p1 <- ggplot(filter(feature, type == "entire protein"), aes(x = id, y = start)) + 
  geom_segment(aes(xend = id, yend = end), size = 2, color = "grey40")
p2 <- geom_segment(data = tango, aes(x = name, xend = name,  y = ifelse(start-4 >= 0, start-4, 0), yend = end + 4, color = median), size = 2)
p3 <- geom_segment(data = filter(feature, type == "Hyphal_reg_CWP"), aes(x = id, y = start, xend = id, yend = end), size = 2, color = "#1f78b4")
p1 + p2 + p3 + coord_flip() + theme_classic() + 
  scale_color_distiller(type = "seq", palette = 17, direction = 1) +
  theme(axis.text.y = element_text(size = 5),
        axis.line.y = element_blank(), axis.ticks.y = element_blank(),
        axis.line.x = element_blank(), axis.ticks.x = element_blank(),
        legend.position = c(0.85,0.25),
        panel.background = element_rect(fill = alpha("lightblue",0.5))) +
  ylim(-2, 4500) + labs(y = "Position in sequence", x = "Sequences") + 
  ggtitle("TANGO predicted aggregation-prone sequences", subtitle = "Hyphal_reg_CWP domain masked")
ggsave("img/20200910-tango-score-segment.png", width = 7, height = 7)
```

```{r, eval=FALSE, include=FALSE}
# read in TANGO output
# tango output are stored as gzipped files for each individual sequence. we need to read in all of the files into a list of tibbles
tango.output.files <- list.files(path = "tango-output", pattern = "*.txt.gz", include.dirs = T)
ltango <- lapply(tango.output.files, function(x) {
  read_tsv(paste0("tango-output/",x), col_types = "icddddd") %>% select(res, aa, Aggregation)
  })
names(ltango) <- gsub(".txt.gz","",tango.output.files)
tango.output <- bind_rows(ltango, .id = "id")
# the id column stores only the refseq_id, without the species name portion. to make sure that the sequences
# are plotted in the desirable order, we do the following
names(genetreeOrder) <- gsub("_[A-Za-z]+$","", genetreeOrder, perl = TRUE)
tango.ordered <- tango.output %>% 
  mutate(new_id = genetreeOrder[id], ordered_id = ordered(new_id, levels = rev(genetreeOrder))) %>% 
  select(ordered_id, res, aa, aggregation = Aggregation)
```

Now we can plot them the same way as we plotted the S/T frequencies:
```{r, eval=FALSE, include=FALSE}
p <- ggplot(tango.ordered, aes(x = ordered_id, y = res)) + geom_tile(aes(fill = aggregation)) + 
  scale_fill_gradient(limits = c(5,100), breaks = c(5,20,50,100), low = "white", high = "yellow", trans = "sqrt") +
  coord_flip() +  theme_classic() +
  theme(axis.text.y = element_text(size = 5),
        axis.line.y = element_blank(), axis.ticks.y = element_blank(),
        axis.line.x = element_blank(), axis.ticks.x = element_blank(),
        legend.position = c(0.9,0.8),
        panel.background = element_rect(fill = alpha("lightblue",0.5))) +
  ylim(1, 4500) + ylab("Position in sequence") + xlab("Sequences") + ggtitle("TANGO predicted aggregation score")
p
ggsave("img/20200811-tango-score-tile.png", p, bg = "transparent", width = 7, height = 7)
```

