---
title: "plot XP_028889033 features"
author: Bin He
date: 2020-10-11
output: 
  html_notebook:
    code_folding: hide
---

## Goal
Plot domain architecture, sequence features and predicted $\beta$-aggregation scores for XP_028889033 (corresponding gene: CJI97_004172, _C. auris_ B11221)

## Data
The following data were extracted from `../homolog-properties/2020-10-31/raw-output` folder
- Serine and threonine frequencies calculated in 100 bp windows with 10 bp increment.
- TANGO predicted aggregation probabilities per amino acid.
The output is saved in `input/XP_028889033-quant-properties.txt`

The domain architecture, including tandem repeat domains, were based on the same folder above plus `../Cauris_polymorphism/output/tandem-repeats` and saved in `input/XP_028889033-domain-features.txt` and `input/XP_028889033-tandem*`

## Plot
```{r message=FALSE}
library(tidyverse)
library(cowplot)
```


```{r read_data}
# quantitative features
dat <- read_tsv("input/XP_028889033-quant-features.txt", col_types = "ffid")
# dat$feature <- factor(dat$feature, labels = c("Ser/Thr frequency","beta-aggregation probability"))

# domain features
domain <- read_tsv("input/XP_028889033-domain-features.txt")
domain$feature <- factor(domain$feature, levels = c("entire protein", "SignalP", "Hyphal_reg_CWP", "Hyr1", "GPI"), labels = c("entire protein", "Signal Peptide", "Hyphal_reg_CWP", "Hyr1 (PF15789)", "GPI-anchor"))

# tandem repeats
tandem <- read_tsv("input/XP_028889033_tandem.tsv", col_types = cols())

# combine tandem repeat domains with domain features
feature <- bind_rows(
  domain %>% filter(!grepl("Hyr1",feature)),
  tandem %>% mutate(feature = paste0("TR", 1:nrow(tandem), " (", period, " aa x ", copyN, ")")) %>% 
    select(feature, start, end)
)

# tandem repeat dividers
tandem.div <- read_tsv("input/XP_028889033_tandem_div.tsv", col_types = cols())
```

```{r plot, fig.width=8}
p1 <- dat %>% filter(feature == "ST_freq") %>% 
  ggplot(aes(x = pos, y = value)) + geom_line(aes(color = type)) +
  xlab("Amino acid position") + ylab("frequency") + ylim(0,0.6) + xlim(1,3100) +
  theme_cowplot() + theme(legend.position = c(0.8, 0.8), legend.text = element_text(size = rel(0.8)))

p2 <- dat %>% filter(feature == "TANGO") %>% 
  ggplot(aes(x = pos, y = value)) + geom_line(size = 0.3) + ylab("TANGO") +
  theme_cowplot() + theme(axis.title.x = element_blank(), axis.text.x = element_blank()) + xlim(1,3100) +
  scale_y_continuous(breaks = c(0,50,100))

feature.col <- c("grey40", "#984ea3", "#377eb8", "#e41a1c", "#fdb863", "#e66101", "#c2a5cf", "#a6dba0", "#008837")
p3 <- ggplot(feature, aes(xmin = start, ymin = 0)) + 
  geom_rect(aes(xmax = end, ymax = 1, fill = feature)) +
  scale_fill_manual(values = feature.col) +
  ylab("") + xlab("") + xlim(1,3100) + ylim(-0.5,1) + labs(fill = "Domain") + guides(fill = guide_legend(nrow = 2)) +
  theme_classic() + theme(legend.position = "top", legend.text = element_text(size = rel(0.8))) +
  theme(axis.text = element_blank(), axis.ticks = element_blank(), axis.line = element_blank())

p4 <- geom_rect(data = tandem.div, aes(xmin = div, ymin = 0, xmax = div + 3, ymax = 1))
# assembling
plot_grid(p3+p4, p2, p1, nrow = 3, align = "v", rel_heights = c(2,1,3))
ggsave("output/r-figure/20210507-XP_028889033-features.png", width = 11, height = 5)
ggsave("output/r-figure/20210507-XP_028889033-ST-freq.png", p1 + theme_classic() + theme(legend.position = "top"), width = 6, height = 2)
```

```{r}
DT::datatable(select(tandem, start, end, period, copyN, consensus_nogap), 
              caption = "Tandem repeats in XP_028889033", 
              options = list(
                deferRender = TRUE,
                scrollX = TRUE,
                scroller = TRUE,
                autoWidth = TRUE,
                pageLength = 5,
                columnDefs = list(list(width = '10%', targets = c(5)))))
```

