---
title: "plot XP_028889033 features"
author: Bin He
date: 2020-10-11
output: html_notebook
---

## Goal
The goal of this script is to plot quantitative features for XP_028889033 (corresponding gene: CJI97_004172, _C. auris_ B11221)

## Data
The following data were extracted from `02-case-studies/output/homolog-properties/2020-07-24/raw-output` folder
- Serine and threonine frequencies calculated in 100 bp windows with 10 bp increment.
- TANGO predicted aggregation probabilities per amino acid.
The output is saved in `protein-quant-properties.txt`

## Plot
```{r}
library(tidyverse)
library(cowplot)
```

```{r plot, fig.width=8}
# quantitative features
dat <- read_tsv("protein-quant-features.txt", col_types = "ffid")
# dat$feature <- factor(dat$feature, labels = c("Ser/Thr frequency","beta-aggregation probability"))
p1 <- dat %>% filter(feature == "ST_freq") %>% 
  ggplot(aes(x = pos, y = value)) + geom_line(aes(color = type)) +
  xlab("Amino acid position") + ylab("frequency") + ylim(0,0.6) + xlim(1,3100) +
  theme_cowplot() + theme(legend.position = c(0.8, 0.8), legend.text = element_text(size = rel(0.8)))

p2 <- dat %>% filter(feature == "TANGO") %>% 
  ggplot(aes(x = pos, y = value)) + geom_line(size = 0.3) + ylab("TANGO Score") +
  theme_cowplot() + theme(axis.title.x = element_blank(), axis.text.x = element_blank()) + xlim(1,3100) +
  scale_y_continuous(breaks = c(0,50,100))
  
# domain features
domain <- read_tsv("protein-domain-features.txt")
domain$feature <- factor(domain$feature, levels = c("entire protein", "SignalP", "Hyphal_reg_CWP", "Hyr1", "GPI"), labels = c("entire protein", "Signal Peptide", "Hyphal_reg_CWP (PF11765)", "Repeated domain in Hyr1 (PF15789)", "GPI-anchor"))
p3 <- ggplot(domain, aes(xmin = start, ymin = 0)) + 
  geom_rect(aes(xmax = end, ymax = 1, fill = feature)) +
  scale_fill_manual(values = c("grey50","#e41a1c", "#377eb8", "#ff7f00", "#984ea3")) +
  ylab("") + xlab("") + xlim(1,3100) + ylim(-0.5,1) + labs(fill = "Domain") + guides(fill = guide_legend(nrow = 2)) +
  theme_classic() + theme(legend.position = "top", legend.text = element_text(size = rel(0.8))) +
  theme(axis.text = element_blank(), axis.ticks = element_blank(), axis.line = element_blank())

# assembling
plot_grid(p3, p2, p1, nrow = 3, align = "v", rel_heights = c(3,2,4))
ggsave("20201011-XP_028889033-features.png", width = 11, height = 5)
ggsave("20201011-XP_028889033-ST-freq.png", p1 + theme_classic() + theme(legend.position = "top"), width = 6, height = 2)
```

